Speed and strength of epidemic intervention

### 2019-07-23 doc/speedstrength.tex
no more merge conflicts?
tweaks

Much research has prioritized estimates of the reproductive number, because it has a threshold value, $\RR=1$ that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts. 
The insight that a case must on average cause at least one new case under good conditions for a disease persist goes back $>100$ years \cite{Ross}, and the idea of averaging by defining a `typical' case was formalized in the 19x0s \cite{Heesterbeek}.
$\RR$ is also of interest because it provides a \emph{prima facie} prediction about the total \emph{size} of an epidemic.

@@

We model disease incidence using the renewal-equation framework \cite{SeeChamp?}. 
This is a simple, flexible framework that can cover a wide range model structures.
In our model, disease incidence at time $t$ is given by:
\begin{equation}
i(t) = \int K(\tau, t) i(t-\tau) d\tau.
\end{equation}
Here, $K$ is the infection kernel describing how infectious we expect an individual infected $\tau$ time units ago to be in the population.
In general, $K$ will depend on population characteristics that may change through time -- notably, the proportion of the population susceptible.
Since we are interested in invasion and control, we will generally neglect changes in $K$ through time, thus we will assume $K\equiv K(\tau)$. 
This means we are neglecting changes in susceptible proportion through time.

@@

Assuming that $K$ doesn't change with time, we write:
\begin{equation}
	K(\tau) = \RR g(\tau),
	\eqlab{speed}
\end{equation}
where $g(\tau)$ is the ``intrinsic'' generation-interval distribution (giving the relative infectiousness of the average individual as a function of time since infection) \cite{Champ}. Since $g$ is a distribution, it integrates to 1, and $\RR$ is thus the integral of $K$.

@@

We generalize this idea by allowing an intervention strategy to reduce $K$ by different proportions over the course of an individual infection. We write the post-intervention kernel:
\begin{equation}
	\hat K(\tau) = K(\tau)/L(\tau), 
\end{equation}
where $L(\tau)$ is the average proportional reduction for an individual infected time $\tau$ ago.

@@

By analogy with the strength-based factorization \eqref{speed}, we can rewrite the Euler-Lotka equation \eqref{euler} as a speed-based factorization:

\begin{equation}
K(\tau) = b(\tau)\exp(r\tau)
\end{equation}

Like $g$, $b$ is a distribution: in this case the initial backward generation interval, which gives the distribution of realized generation times when the disease spreads exponentially \cite{Champ}.

Now imagine an intervention that reduces transmission at a constant hazard rate $\phi$ across the disease generation, for example by identifying and isolating infectious individuals.
We then have:
\begin{equation}
	\hat K(\tau) = K(\tau)\exp(-\phi\tau) = b(\tau)\exp((r-\phi)\tau)
\end{equation}
Since $b$ is a distribution, the reduction needed to prevent invasion (or to eliminate disease)  is exactly $\phi=r$. We call $\phi$ the ``speed'' of the intervention; transmission is interrupted when the speed of the intervention is faster than the speed of spread.

We generalize this idea by allowing the hazard rate $h(\tau)$ at which $K$ is reduced to vary through time, thus:
\begin{equation}
	\hat K(\tau) = K(\tau) \exp(-\int_0^\tau h(\sigma) d\sigma)
\end{equation}

We define the speed of a general intervention $L(\tau)$ \eqref{Ldef} to be $\phi = r - \hat r$. We can then show that $\phi$ is a (sort of) mean satisfying:
\begin{equation}
	1 = \left\langle \frac{\exp(\phi \tau) }{\exp(\int_0^\tau h(\sigma) d\sigma)} \right\rangle_{b(\tau)}
\end{equation}
The speed $\phi$ is a mean of the hazard $h$ in the sense that an increase (or decrease) in $h$ produces the same sign of change in $\phi$, and if $h$ is constant across the generation then $\phi=h$.

\section{Examples}

@@

>>>>

Much research has prioritized estimates of the reproductive number, because it has a threshold value, $\RR=1$ that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts. 
The insight that a case must on average cause at least one new case under good conditions for a disease persist goes back $>100$ years \citep{ross1911prevention};
the idea of averaging by defining a `typical' case was formalized in the 1990s \citep{diekmann1990definition}.
$\RR$ is also of interest because it provides a \emph{prima facie} prediction about the total \emph{size} of an epidemic \citep{anderson1991infectious, ma2006generality, miller2012note}.

@@

\swp{edited to make the notation slightly clearer}
We model disease incidence using the renewal-equation framework \citep{champredon2015intrinsic}. 
This is a simple, flexible framework that can cover a wide range model structures \citep{Champredon2018equivalence}.
In our model, disease incidence at time $t$ is given by:
\begin{equation}
i(t) = \int K(\tau, t) i(t-\tau) d\tau.
\end{equation}
Here, $K(\tau, t)$ is the infection kernel describing how infectious we expect an individual infected $\tau$ time units ago to be in the population.
In general, $K(\tau, t)$ will depend on population characteristics that may change through time $t$ -- notably, the proportion of the population susceptible.
Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time, thus we will assume $K(\tau, t) \equiv K(\tau)$. 
This means we are neglecting changes in susceptible proportion through time;
under this assumption, the renewal equation is equivalent to the Von Foerster equations, outlined in \cite{fraser2004factors}.

@@

Assuming that the infection kernel $K$ doesn't change with time, we write:
\begin{equation}
	K(\tau) = \RR g(\tau),
	\eqlab{strength}
\end{equation}
where $g(\tau)$ is the ``intrinsic'' generation-interval distribution (giving the relative infectiousness of the average individual as a function of time since infection) \citep{champredon2015intrinsic}. 
Since $g$ is a distribution, it integrates to 1, and the reproduction number $\RR$ is thus the integral of $K$.

@@

We generalize this idea by allowing an intervention strategy to reduce $K$ by different proportions over the course of an individual infection. We write the post-intervention kernel:
\begin{equation}
	\hat K(\tau) = K(\tau)/L(\tau), 
\end{equation}
where $L(\tau)$ is the average proportional reduction for an individual infected time $\tau$ ago.
Then, we define the post-intervention reproductive number with the following integral:
\begin{equation}
	\Rhat = \int K(\tau) d\tau
\end{equation}
This framework generalizes the work of \cite{fraser2004factors} who made parameteric assumptions about the shape of $1/L(\tau)$.

@@

The Euler-Lokta equation allows us to calculate the initial exponential growth rate $r$ of an epidemic given an infection kernel $K$:
\begin{equation}
	1 = \int K(\tau) \exp(-r\tau) d\tau
	\eqlab{euler}
\end{equation}
By analogy with the strength-based factorization \eqref{strength}, we can rewrite the Euler-Lotka equation \eqref{euler} as a speed-based factorization:
\begin{equation}
K(\tau) = b(\tau)\exp(r\tau)
\end{equation}
Like $g$, $b$ is a distribution: in this case the initial backward generation interval, which gives the distribution of realized generation times (measured from the infectee's point of view) when the disease spreads exponentially \citep{champredon2015intrinsic, britton2019estimation}.

Now imagine an intervention that reduces transmission at a constant hazard rate $\phi$ across the disease generation, for example by identifying and isolating infectious individuals.
We then have:
\begin{equation}
	\hat K(\tau) = K(\tau)\exp(-\phi\tau) = b(\tau)\exp((r-\phi)\tau)
\end{equation}
Since $b$ is a distribution (which integrates to 1), the reduction needed to prevent invasion (or to eliminate disease)  is exactly $\phi=r$. 
We call $\phi$ the ``speed'' of the intervention; transmission is interrupted when the speed of the intervention is faster than the speed of spread.

We generalize this idea by allowing the hazard rate $h(\tau)$ at which $K$ is reduced to vary through time, thus:
\begin{equation}
	\hat K(\tau) = K(\tau) \exp\left(-\int_0^\tau h(\sigma) d\sigma\right)
\end{equation}
The associated post-intervention epidemic speed $\hat r$ is given by:
\begin{equation}
	1 = \int \hat K(\tau) \exp(-\hat r\tau) d\tau.	
\end{equation}
We define the speed of a general intervention to be $\phi = r - \hat r$. 
We can then show that $\phi$ is a (sort of) mean satisfying:
\begin{equation}
	1 = \left\langle \frac{\exp(\phi \tau) }{\exp\left(\int_0^\tau h(\sigma) d\sigma\right)} \right\rangle_{b(\tau)}
\end{equation}
The speed $\phi$ is a mean of the hazard $h$ in the sense that an increase (or decrease) in $h$ produces the same sign of change in $\phi$, and if $h$ is constant across the generation then $\phi=h$.

\section{Results}

In this section, we use both strength- and speed-based decompositions to compare different intervention strategies for HIV. In particular, we compare how the amount of early vs. late transmission affect each intervention strategies.

In order to

@@

\bibliography{speedstrength}
